Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-23 boettiger.tex
submitted to JRSI
corrected transcritical bifurcation statement

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
they are characterized by a steady approach towards the alternative stable state, 
such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

\section{Acknowledgments}
We would like to thank S. Schreiber, M. Holyoak, M. Baskett, and A. Perkins for comments on the manuscript. 
This research was supported by funding from NSF Grant EF 0742674 
and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308. 
Data and code are available at \href{https://github.com/cboettig/earlywarning}{ https://github.com/cboettig/earlywarning}.

>>>>

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome.% --
%they are characterized by a steady approach towards the alternative stable state, 
%such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

\section{Acknowledgments}
We would like to thank S. Schreiber, M. Holyoak, M. Baskett, A. Perkins and N. Ross for comments on the manuscript. 
This research was supported by funding from NSF Grant EF 0742674 
and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308. 
Data and code are available at \href{https://github.com/cboettig/earlywarning}{ https://github.com/cboettig/earlywarning}.
